Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections

### 2023-05-22 jss/cheem_paper.tex
-fix app link -minor lint&wordsmithing -motiviating letter dates
edits through-out

\Abstract{
The increased predictive power of machine learning models comes at the
cost of increased complexity and loss of interpretability, particularly
in comparison to parametric statistical models. This trade-off has led
to the emergence of eXplainable AI (XAI) which provides methods, such as
local explanations (LEs) and local variable attributions (LVAs), to shed
light on how a model use predictors to arrive at a prediction. These
provide a point estimate of the linear variable importance in the
vicinity of a single observation. However, LVAs tend not to effectively
handle association between predictors. To understand how the interaction
between predictors affects the variable importance estimate, we can
convert LVAs into linear projections and use the radial tour. This is
also useful for learning how a model has made a mistake, or the effect
of outliers, or clustering of observations. The approach is llustrated
with examples from categorical (penguin species, chocolate types) and
quantitative (soccer/football salaries, house prices) response models.
The methods are implemented in the R package cheem, available on CRAN.
}

@@

\usepackage{amsmath}

@@

There are different reasons and purposes for fitting a model. According
to the taxonomies of \citet{breiman_statistical_2001} and
\citet{shmueli_explain_2010}, it can be useful to group models into two
types: explanatory and predictive. Explanatory modeling is used for
inferential purposes, while predictive modeling focuses solely on the
performance of an objective function. The intended use of the model has
important implications for its selection and development.
Interpretability is critical in explanatory modeling to draw meaningful
inferential conclusions, such as which variables most contribute to a
prediction or whether some observations are less well fit.
Interpretability becomes more difficult when the model is nonlinear.
Nonlinear models occur in statistical models with polynomial or
interaction terms between quantitative predictors, and almost all
computational models such as random forests, support-vector machines, or
neural networks
\citep[e.g.][]{breiman_random_2001, boser_training_1992, anderson_introduction_1995}.

In linear models interpretation of the importance of variables is
relatively straight forward, one adjusts for the covariance of multiple
variables when examining the relationship with the response. The
interpretation is valid for the full domain of the predictors. In
nonlinear models one needs to consider the model in small neighborhoods
of the domain to make any assessment of variable importance. Even though
this is difficult, it is especially important to interpret model fits as
we become more dependent on nonlinear models for routine aspects of life
to avoid issues described in \citet{stahl-ethics}. Understanding how
nonlinear models behave when usage extrapolates outside the domain of
predictors, either in sub-spaces where few samples were provided in the
training set, or extending outside the domain. It is especially
important because nonlinear models can vary wildly and predictions can
be dramatically wrong in these areas.

@@

For multivariate data visualization, a \emph{tour}
\citep{asimov_grand_1985, buja_grand_1986, lee_state_2021} of linear
data projections onto a lower-dimensional space, could be an element of
XAI, complementing LVAs. Applying tours to model interpretation is
recommended in \citet{wickham_visualizing_2015} primarily to examine the
fitted model in the space of the data. \citet{cook_interactive_2007}
describe the use of tours for exploring classification boundaries and
model diagnostics
\citep{Caragea2008, lee_pptree_2013, da_silva_projection_2021}. There
are various types of tours. In a \emph{manual} or radial tour
\citep{cook_manual_1997, spyrison_spinifex_2020}, the path of linear
projections is defined by changing the contribution of a selected
variable. We propose to use this to scrutinize the LVAs. This approach
could be considered to be a counter-factual, what-if analysis, such as
\emph{ceteris paribus} (``other things held constant'') profiles
\citep{biecek_ceterisparibus_2020}.

The remainder of this paper is organized as follows. Section
\ref{sec:explanations} covers the background of the LEs and the
traditional visuals produced. Section \ref{sec:tour} explains the tours
and particularly the radial manual tour. Section \ref{sec:cheemviewer}
discusses the visual layout in the graphical user interface and how it
facilitates analysis, data pre-processing, and package infrastructure.
Illustrations are provided in Section \ref{sec:casestudies} for a range
of supervised learning tasks with categorical and quantitative response
variables. These show how the LVAs can be used to get an overview of the
model's use of predictors, and to investigate errors in the model
predictions. Section \ref{sec:cheemdiscussion} concludes with a summary
of the insights gained. The methods are implemented in the \textbf{R}
package \textbf{cheem}.

@@

We will be focusing on SHAP values in this this paper, but the approach
is applicable for any method used to calculate the LVAs. SHAP calculates
the variable contributions of one observation by examining the effect of
other variables on the predictions. The term ``SHAP'' refers to
\citet{shapley_value_1953}'s method to evaluate an individual's
contribution in cooperative games by assessing this player's performance
in the presence or absence of other players.
\citet{strumbelj_efficient_2010} introduced SHAP for LEs in machine
learning models. Variable importance can depend on the sequence in which
variables are entered into the model fitting process, thus for any
sequence we get a set of variable contribution values for a single
observation. These values will add up to the difference between the
fitted value for the observation, and the average fitted value for all
observations. Using all possible sequence, or permutation, gives
multiple values for each variable, which are averaged to get the SHAP
value for an observation. It can be helpful to standardize variables
prior to computing SHAP values, if they have been measured on different
scales.

The approach is related to partial dependence plots (see for example,
chapter 8 of \citet{molnar2022}), used to explain the effect of a
variable by predicting the response for a range of values on this
variable after fixing the value of all other variables to their mean.
Though partial dependence plots are a global approximation of the
variable importance, while SHAP is specific to one observation.

@@

We use 2020 season FIFA data \citep{leone_fifa_2020} to illustrate SHAP
following the procedures described in \citet{biecek_explanatory_2021}.
There are 5000 observations of nine predictor variables measuring
players' skills and one response variable, wages (in euros). A random
forest model is fit regressing players' wages on the skill variables. In
this illustration in Figure \ref{fig:shapdistrbd} the SHAP values are
compared for a star offensive player (L. Messi) and a prominent
defensive player (V. van Dijk). We are interested in knowing how the
skill variables locally contribute to the wage prediction of each
player. A difference in the attribution of the variable importance
across the two positions of the players can be expected. This would be
interpreted as how a player's salary depends on which combination of
skills. Panel (a) is a version of a breakdown plot
\citep{gosiewska_ibreakdown_2019} where just three sequences of
variables are shown, for two observations. A breakdown plot shows the
absolute values of the variable attribution for an observation, usually
sorted from highest value to the lowest. There is no scale on the
horizontal axis here because values are considered relative to each
other. Here we can see how the variable contribution can change
depending on sequence, relative to both players. (Note that the order of
the variables is different in each plot, because they have been sorted
by biggest average contribution across both players.) For all sequences,
and for both players \texttt{reaction} has the strongest contribution,
with perhaps more importance for the defensive player. Then it differs
by player: for Messi \texttt{offense} and \texttt{movement} have the
strongest contributions, and for van Dijk it is \texttt{defense} and
\texttt{power}, regardless of the variable sequence.

Panel (b) shows the differences in the player's median values (large
dots) for 25 such sequences (tick marks). We can see that the wage
predictions for the two players come from different combinations of
skills sets, as might be expected for players who's value on the team
depends on their offensive or defensive prowess. It is also interesting
to see from the distribution of values across the different sequence of
variables, that there is some multimodality. For example, look at the
SHAP values for \texttt{reaction} for Messi, and in some sequences
reaction has a much lower contribution than others. This suggests that
other variables (\texttt{offense}, \texttt{movement} probably) can
substitute for \texttt{reaction} in the wage prediction.

@@

There are numerous R packages currently available which provide
functions for computing SHAP values, including \texttt{fastshap}
\citep{fastshap}, \texttt{kernelshap} \citep{kernelshap}, \texttt{shapr}
\citep{shapr}, \texttt{shapviz} \citep{shapviz}, \texttt{PPtreeregViz}
\citep{PPtreeregViz}, \texttt{ExplainPrediction}
\citep{ExplainPrediction}, \texttt{flashlight} \citep{flashlight} and
the package \texttt{DALEX} has many resources \citep{biecek_dalex_2018}.
There are many more packages only available through Github, like
\texttt{treeshap} \citep{kominsarczyk_treeshap_2021} that is used for
this work. \citet{molnar2022} provides good explanations of the
different methods and how to apply them to different models.

@@

The manual tour \citep{cook_manual_1997} defines its path by changing a
selected variable's contribution to a basis to allow the variable to
contribute more or less to the projection. The requirement constrains
the contribution of all other variables that a basis needs to be
orthonormal (column correspond to vectors, with unit length, and
orthogonal to each other). The manual tour is primarily used to assess
the importance of a variable to structure visible in a projection. It
also lends itself to pre-computation queued in advance or computed
on-the-fly for human-in-the-loop analysis
\citep{karwowski_international_2006}.

@@

\caption{The radial tour allows the user to remove a variable from a projection, to examine the importance of this variable to structure in the plot. Here we have a 1D projection of the penguins data displayed as a density plot. The line segments on the bottom correspond to the coefficients of the variables making up the projection. The structure in the plot is bimodality (left), and the importance of the variable \textsf{bd} is being explored. As this variable contribution is reduced in the plot (middle, right) we can see that the bimodality decreases. Thus \textsf{bd} is an important variable contributing to the bimodal structure.}\label{fig:radialtour}
\end{figure}
\end{CodeChunk}

@@

The visualization is composed of the first two principal components of
the data (left) and the attribution (middle) spaces. These single 2D
projections will not reveal all of the structure of higher-dimensional
space, but they are helpful visual summaries. In addition, a plot of the
observed against predicted response values is also provided (Figures
\ref{fig:classificationcase}b, \ref{fig:regressioncase}a) to help
identify observations poorly predicted by the model. For classification
tasks, color indicates the predicted class, and misclassified
observations are circled in red. Linked brushing between the plots is
provided, and a tabular display of selected points helps to facilitate
exploration of the spaces and the model (shown in Figures
\ref{fig:regressioncase}d).

While the comparison of these spaces is interesting, the primary purpose
of the global view is to enable the selection of particular observations
to explore in detail. We have designed it to enable a comparison between
an observation that is interesting is some way, perhaps misclassified,
or poorly predicted, relative to an observation with similar predictor
values but more expected prediction. For brevity, we call the
interesting observation the primary investigation (PI) and the other is
the comparison investigation (CI). These observations are highlighted as
asterisk and \(\times\), respectively.

@@

There are two plots in this part of the interface. The first (Figures
\ref{fig:classificationcase}e and \ref{fig:regressioncase}e) is a
display of the SHAP values for all observations. This will generally
give the global view of variables important for the fit as a whole, but
it will also highlight observations which have different patterns. The
second plot is the radial tour, which for classification is a density
plot of a 1D projection (Figure \ref{fig:classificationcase}f), and for
regression are scatterplots of the observed response values, and
residuals, against a 1D projection (Figure \ref{fig:regressioncase}f).

The LVAs for all observations are normalized (sum of squares equals 1),
and thus, the relative importance of variables can be compared across
all observations. These are depicted as a vertical parallel coordinate
plot \citep{ocagne_coordonnees_1885}. (The SHAP values of the PI and CI
are shown as dashed and dotted lines, respectively.) One should obtain a
sense of the overall importance of variables from this plot. The more
important variables will have larger values, and in the case of
classification tasks variables which have different magnitude for
different classes are more globally important. For example, Figure
\ref{fig:classificationcase}e suggests that \texttt{bl} is important for
distinguishing the green class from the other two. For regression, one
might generally observe which variables have low values for all
observations (not important). For example, \texttt{BMI} and \texttt{pwr}
in Figure \ref{fig:regressioncase}e, have a range of high and low values
(e.g., \texttt{off}, \texttt{def}), suggesting they are important for
some observations and not important for others.

A bar chart is overlaid to represent the projection shown in the radial
tour at right. It starts from the SHAP values of the PI, but if the user
changes the projection the length of these bars will reflect this
change. (The PI is interactively selected by clicking on a point in the
global view). By scaling the SHAP value it becomes an (attribution)
projection.

@@

\caption[Overview of the cheem viewer for regression tasks (quantitative response) and illustration of interactive variables]{Overview of the cheem viewer for regression tasks (quantitative response) and illustration of interactive variables. Panel (a) PCA of the data- and attributions- spaces and the (b) residual plot, predictions by observed values. Four selected points are highlighted in the PC spaces and tabularly displayed. Coloring on a statistic (c) highlights structure organized in the attribution space. Interactive tabular display (d) populates when observations are selected. Contribution of the 1D basis affecting the horizontal position (e) parallel coordinate display of the variable attribution from all observations, and horizontal bars show the contribution to the current basis. Regression projection (f) uses the same horizontal projection and fixes the vertical positions to the observed $y$ and residuals (middle and right).}\label{fig:regressioncase}
\end{figure}
\end{CodeChunk}

@@

A tooltip displays observation number/name and classification
information while the cursor hovers over a point. Linked brushing allows
the selection of points (left click and drag) where those points will be
highlighted across plots (Figure \ref{fig:classificationcase}a \& b).
The information corresponding to the selected points is populated on a
dynamic table (Figure \ref{fig:classificationcase}d). These interactions
aid exploration of the spaces and, finally, the identification of
primary and comparison observations.

@@

\begin{itemize}
\tightlist
\item
  \textbf{Data:} predictors and response are unscaled complete numerical
  matrix. Most models and local explanations are scale-invariant. Keep
  normality assumptions of the model in mind.
\item
  \textbf{Model:} any model and compatible explanation could be explored
  with this method. Currently, random forest models are applied via the
  package \textbf{randomForest} \citep{liaw_classification_2002},
  compatibility tree SHAP. Modest hyperparameters are used, namely: 125
  trees, the number of variables at each split, mtry = \(\sqrt{p}\) or
  \(p/3\) for classification and regression, and minimum size of
  terminal nodes \(max(1, n/500)\) or \(max(5, n/500)\) for
  classification and regression.
\item
  \textbf{Local explanation:} Tree SHAP is calculated for \emph{each}
  observation using the package \textbf{treeshap}
  \citep{kominsarczyk_treeshap_2021}. We opt to find the attribution of
  each observation in the training data and not fit to fit variable
  interactions.
\item
  \textbf{Cheem viewer:} after the model and full explanation space are
  calculated, each variable is scaled by standard deviations away from
  the mean to achieve common support for visuals. Statistics for mapping
  to color are computed on the scaled spaces.
\end{itemize}

The time to preprocess the data will vary significantly with the
complexity of the model and the LE. For reference, the FIFA data
contained 5000 observations of nine explanatory variables took 2.5
seconds to fit a random forest model of modest hyperparameters.
Extracting the tree SHAP values of each observation took 270 seconds
total. PCA and statistics of the variables and attributions took 2.8
seconds. These run times were from a non-parallelized session on a
modern laptop, but suffice to say that most of the time will be spent on
the LVA. An increase in model complexity or data dimensionality will
quickly become an obstacle. Its reduced computational complexity makes
tree SHAP an excellent candidate to start. Alternatively, the package
\textbf{fastshap} claims extremely low run times, attributed to fewer
calls to the prediction function, partial implementation in C++, and
efficient use of logical subsetting, \citet{greenwell_fastshap_2020}.

@@

There is more separation visible in the attribution space than in the
data space, as would be expected. The predicted vs observed plot reveals
a handful of misclassified observations. A Gentoo that has been wrongly
labeled as a Chinstrap is selected for illustration. The PI is a
misclassified point (represented by the asterisk in the global view and
a dashed vertical line in the tour view). The CI is a correctly
classified point (represented by an \(\times\) and a vertical dotted
line).

The radial tour starts from the attribution projection of the
misclassified observation (b, left). The important variables identified
by SHAP in the (wrong) prediction for this observation are mostly
\texttt{bl} and \texttt{bd} with small contributions of \texttt{fl} and
\texttt{bm}. This projection is a view where the Gentoo (purple) looks
much more likely for this observation than Chinstrap. That is, this
combination of variables is not particularly useful because the PI looks
very much like other Gentoo penguins. The radial tour is used to vary
the contribution of flipper length (\texttt{fl}) to explore this. (In
our exploration, this was the third variable explored. It is typically
helpful to explore the variables with more significant contributions,
here \texttt{bl} and \texttt{bd}. Still, when doing this, nothing was
revealed about how the PI differed from other Gentoos). On varying
\texttt{fl} as it contributes increasingly to the projection (b, right),
more and more, this penguin looks like a Chinstrap. This suggests that
\texttt{fl} should be considered an important variable for explaining
the (wrong) prediction.

@@

\caption[Examining the LVA for a PI which is dark (orange) chocolate incorrectly predicted to be milk (green)]{Examining the LVA for a PI which is dark (orange) chocolate incorrectly predicted to be milk (green). From the attribution projection, this chocolate correctly looks more like dark than milk, which suggests that the LVA does not help understand the prediction for this observation. So, the contribution of Sugar is varied---reducing it corresponds primarily with an increased magnitude from Fiber. When Sugar is zero, Fiber contributes strongly towards the left. In this view, the PI is closer to the bulk of the milk chocolates, suggesting that the prediction put a lot of importance on Fiber. This chocolate is a rare dark chocolate without any Fiber leading to it being mistaken for a milk chocolate. (A video of the tour animation can be found at https://vimeo.com/666431143.)}\label{fig:casechocolates}
\end{figure}
\end{CodeChunk}

@@

Figure \ref{fig:caseames} selects the house sale 74, a sizable under
prediction with an enormous Lot Area contribution. The CI has a similar
predicted price though the prediction was accurate and gives almost no
attribution to lot size. The attribution projection places observations
with high Living Areas to the right. The contribution of Living Area
contrasts the contribution of this variable. As the contribution of Lot
Area decreases, the predictive power decreases for the PI, while the CI
remains stationary. This large importance in the Living Area is
relatively uncommon. Boosting tree models may be more resilient to such
an under-prediction as they would up-weighting this residual and force
its inclusion in the final model.

@@

To address this challenge in this paper provides a technique that builds
on LEs to explore the variable importance local to an observation. The
LVA is converted into an attribution projection from which variable
contributions are varied using a radial tour. Several diagnostic plots
are provided to assist with understanding the sensitivity of the
prediction to particular variables. A global view shows the data space,
explanation space, and residual plot. The user can interactively select
observations to compare, contrast, and study further. Then the radial
tour is used to explore the variable sensitivity identified by the
attribution projection.

>>>>

\Abstract{
The increased predictive power of machine learning models comes at the
cost of increased complexity and loss of interpretability, particularly
in comparison to parametric statistical models. This trade-off has led
to the emergence of eXplainable AI (XAI) which provides methods, such as
local explanations (LEs) and local variable attributions (LVAs), to shed
light on how a model use predictors to arrive at a prediction. These
provide a point estimate of the linear variable importance in the
vicinity of a single observation. However, LVAs tend not to effectively
handle association between predictors. To understand how the interaction
between predictors affects the variable importance estimate, we can
convert LVAs into linear projections and use the radial tour. This is
also useful for learning how a model has made a mistake, or the effect
of outliers, or the clustering of observations. The approach is
illustrated with examples from categorical (penguin species, chocolate
types) and quantitative (soccer/football salaries, house prices)
response models. The methods are implemented in the R package cheem,
available on CRAN.
}

@@

\usepackage{amsmath} \usepackage{datetime}

@@

There are different reasons and purposes for fitting a model. According
to the taxonomies of \citet{breiman_statistical_2001} and
\citet{shmueli_explain_2010}, it can be useful to group models into two
types: explanatory and predictive. Explanatory modeling is used for
inferential purposes, while predictive modeling focuses solely on the
performance of an objective function. The intended use of the model has
important implications for its selection and development.
Interpretability is critical in explanatory modeling to draw meaningful
inferential conclusions, such as which variables most contribute to a
prediction or whether some observations are less well fit.
Interpretability becomes more difficult when the model is nonlinear.
Nonlinear models occur in statistical models with polynomial or
interaction terms between quantitative predictors, and almost all
computational models such as random forests, support vector machines, or
neural networks
\citep[e.g.][]{breiman_random_2001, boser_training_1992, anderson_introduction_1995}.

In linear models interpretation of the importance of variables is
relatively straightforward, one adjusts for the covariance of multiple
variables when examining the relationship with the response. The
interpretation is valid for the full domain of the predictors. In
nonlinear models, one needs to consider the model in small neighborhoods
of the domain to make any assessment of variable importance. Even though
this is difficult, it is especially important to interpret model fits as
we become more dependent on nonlinear models for routine aspects of life
to avoid issues described in \citet{stahl-ethics}. Understanding how
nonlinear models behave when usage extrapolates outside the domain of
predictors, either in sub-spaces where few samples were provided in the
training set, or extending outside the domain. It is especially
important because nonlinear models can vary wildly and predictions can
be dramatically wrong in these areas.

@@

For multivariate data visualization, a \emph{tour}
\citep{asimov_grand_1985, buja_grand_1986, lee_state_2021} of linear
data projections onto a lower-dimensional space, could be an element of
XAI, complementing LVAs. Applying tours to model interpretation is
recommended by \citet{wickham_visualizing_2015} primarily to examine the
fitted model in the space of the data. \citet{cook_interactive_2007}
describe the use of tours for exploring classification boundaries and
model diagnostics
\citep{Caragea2008, lee_pptree_2013, da_silva_projection_2021}. There
are various types of tours. In a \emph{manual} or radial tour
\citep{cook_manual_1997, spyrison_spinifex_2020}, the path of linear
projections is defined by changing the contribution of a selected
variable. We propose to use this to scrutinize the LVAs. This approach
could be considered to be a counter-factual, what-if analysis, such as
\emph{ceteris paribus} (``other things held constant'') profiles
\citep{biecek_ceterisparibus_2020}.

The remainder of this paper is organized as follows. Section
\ref{sec:explanations} covers the background of the LEs and the
traditional visuals produced. Section \ref{sec:tour} explains the tours
and particularly the radial manual tour. Section \ref{sec:cheemviewer}
discusses the visual layout in the graphical user interface and how it
facilitates analysis, data pre-processing, and package infrastructure.
Illustrations are provided in Section \ref{sec:casestudies} for a range
of supervised learning tasks with categorical and quantitative response
variables. These show how the LVAs can be used to get an overview of the
model's use of predictors and to investigate errors in the model
predictions. Section \ref{sec:cheemdiscussion} concludes with a summary
of the insights gained. The methods are implemented in the \textbf{R}
package \textbf{cheem}.

@@

We will be focusing on SHAP values in this paper, but the approach is
applicable to any method used to calculate the LVAs. SHAP calculates the
variable contributions of one observation by examining the effect of
other variables on the predictions. The term ``SHAP'' refers to
\citet{shapley_value_1953}'s method to evaluate an individual's
contribution in cooperative games by assessing this player's performance
in the presence or absence of other players.
\citet{strumbelj_efficient_2010} introduced SHAP for LEs in machine
learning models. Variable importance can depend on the sequence in which
variables are entered into the model fitting process, thus for any
sequence we get a set of variable contribution values for a single
observation. These values will add up to the difference between the
fitted value for the observation, and the average fitted value for all
observations. Using all possible sequences, or permutations, gives
multiple values for each variable, which are averaged to get the SHAP
value for an observation. It can be helpful to standardize variables
prior to computing SHAP values if they have been measured on different
scales.

The approach is related to partial dependence plots (for example see
chapter 8 of \citet{molnar2022}), used to explain the effect of a
variable by predicting the response for a range of values on this
variable after fixing the value of all other variables to their mean.
Though partial dependence plots are a global approximation of the
variable importance, while SHAP is specific to one observation.

@@

We use 2020 season FIFA data \citep{leone_fifa_2020} to illustrate SHAP
following the procedures described in \citet{biecek_explanatory_2021}.
There are 5000 observations of nine predictor variables measuring
players' skills and one response variable, wages (in euros). A random
forest model is fit regressing players' wages on the skill variables. In
this illustration in Figure \ref{fig:shapdistrbd} the SHAP values are
compared for a star offensive player (L. Messi) and a prominent
defensive player (V. van Dijk). We are interested in knowing how the
skill variables locally contribute to the wage prediction of each
player. A difference in the attribution of the variable importance
across the two positions of the players can be expected. This would be
interpreted as how a player's salary depends on which combination of
skills. Panel (a) is a version of a breakdown plot
\citep{gosiewska_ibreakdown_2019} where just three sequences of
variables are shown, for two observations. A breakdown plot shows the
absolute values of the variable attribution for an observation, usually
sorted from the highest value to the lowest. There is no scale on the
horizontal axis here because values are considered relative to each
other. Here we can see how the variable contribution can change
depending on sequence, relative to both players. (Note that the order of
the variables is different in each plot because they have been sorted by
the biggest average contribution across both players.) For all
sequences, and for both players \texttt{reaction} has the strongest
contribution, with perhaps more importance for the defensive player.
Then it differs by player: for Messi \texttt{offense} and
\texttt{movement} have the strongest contributions, and for van Dijk it
is \texttt{defense} and \texttt{power}, regardless of the variable
sequence.

Panel (b) shows the differences in the player's median values (large
dots) for 25 such sequences (tick marks). We can see that the wage
predictions for the two players come from different combinations of
skill sets, as might be expected for players whose value on the team
depends on their offensive or defensive prowess. It is also interesting
to see from the distribution of values across the different sequences of
variables, that there is some multimodality. For example, look at the
SHAP values for \texttt{reaction} for Messi, and in some sequences,
reaction has a much lower contribution than others. This suggests that
other variables (\texttt{offense}, \texttt{movement} probably) can
substitute for \texttt{reaction} in the wage prediction.

@@

There are numerous R packages currently available that provide functions
for computing SHAP values, including \texttt{fastshap} \citep{fastshap},
\texttt{kernelshap} \citep{kernelshap}, \texttt{shapr} \citep{shapr},
\texttt{shapviz} \citep{shapviz}, \texttt{PPtreeregViz}
\citep{PPtreeregViz}, \texttt{ExplainPrediction}
\citep{ExplainPrediction}, \texttt{flashlight} \citep{flashlight}, and
the package \texttt{DALEX} has many resources \citep{biecek_dalex_2018}.
There are many more packages only available through Github, like
\texttt{treeshap} \citep{kominsarczyk_treeshap_2021} that is used for
this work. \citet{molnar2022} provides good explanations of the
different methods and how to apply them to different models.

@@

The manual tour \citep{cook_manual_1997} defines its path by changing a
selected variable's contribution to a basis to allow the variable to
contribute more or less to the projection. The requirement constrains
the contribution of all other variables that a basis needs to be
orthonormal (columns correspond to vectors, with unit length, and
orthogonal to each other). The manual tour is primarily used to assess
the importance of a variable to the structure visible in a projection.
It also lends itself to pre-computation queued in advance or computed on
the fly for human-in-the-loop analysis
\citep{karwowski_international_2006}.

@@

\caption{The radial tour allows the user to remove a variable from a projection, to examine the importance of this variable to the structure in the plot. Here we have a 1D projection of the penguins data displayed as a density plot. The line segments on the bottom correspond to the coefficients of the variables making up the projection. The structure in the plot is bimodality (left), and the importance of the variable \textsf{bd} is being explored. As this variable contribution is reduced in the plot (middle, right) we can see that the bimodality decreases. Thus \textsf{bd} is an important variable contributing to the bimodal structure.}\label{fig:radialtour}
\end{figure}
\end{CodeChunk}

@@

The visualization is composed of the first two principal components of
the data (left) and the attribution (middle) spaces. These single 2D
projections will not reveal all of the structure of higher-dimensional
space, but they are helpful visual summaries. In addition, a plot of the
observed against predicted response values is also provided (Figures
\ref{fig:classificationcase}b, \ref{fig:regressioncase}a) to help
identify observations poorly predicted by the model. For classification
tasks, color indicates the predicted class and misclassified
observations are circled in red. Linked brushing between the plots is
provided, and a tabular display of selected points helps to facilitate
the exploration of the spaces and the model (shown in Figures
\ref{fig:regressioncase}d).

While the comparison of these spaces is interesting, the primary purpose
of the global view is to enable the selection of particular observations
to explore in detail. We have designed it to enable a comparison between
an observation that is interesting in some way, perhaps misclassified,
or poorly predicted, relative to an observation with similar predictor
values but a more expected prediction. For brevity, we call the
interesting observation the primary investigation (PI), and the other is
the comparison investigation (CI). These observations are highlighted as
an asterisk and \(\times\), respectively.

@@

There are two plots in this part of the interface. The first (Figures
\ref{fig:classificationcase}e and \ref{fig:regressioncase}e) is a
display of the SHAP values for all observations. This will generally
give the global view of variables important for the fit as a whole, but
it will also highlight observations that have different patterns. The
second plot is the radial tour, which for classification is a density
plot of a 1D projection (Figure \ref{fig:classificationcase}f), and for
regression are scatterplots of the observed response values, and
residuals, against a 1D projection (Figure \ref{fig:regressioncase}f).

The LVAs for all observations are normalized (sum of squares equals 1),
and thus, the relative importance of variables can be compared across
all observations. These are depicted as a vertical parallel coordinate
plot \citep{ocagne_coordonnees_1885}. (The SHAP values of the PI and CI
are shown as dashed and dotted lines, respectively.) One should obtain a
sense of the overall importance of variables from this plot. The more
important variables will have larger values, and in the case of
classification tasks variables that have different magnitudes for
different classes are more globally important. For example, Figure
\ref{fig:classificationcase}e suggests that \texttt{bl} is important for
distinguishing the green class from the other two. For regression, one
might generally observe which variables have low values for all
observations (not important). For example, \texttt{BMI} and \texttt{pwr}
in Figure \ref{fig:regressioncase}e, have a range of high and low values
(e.g., \texttt{off}, \texttt{def}), suggesting they are important for
some observations and not important for others.

A bar chart is overlaid to represent the projection shown in the radial
tour on the right. It starts from the SHAP values of the PI, but if the
user changes the projection the length of these bars will reflect this
change. (The PI is interactively selected by clicking on a point in the
global view). By scaling the SHAP value it becomes an (attribution)
projection.

@@

\caption[Overview of the cheem viewer for regression tasks (quantitative response) and illustration of interactive variables]{Overview of the cheem viewer for regression tasks (quantitative response) and illustration of interactive variables. Panel (a) PCA of the data- and attributions- spaces and the (b) residual plot, predictions by observed values. Four selected points are highlighted in the PC spaces and tabularly displayed. Coloring on a statistic (c) highlights the structure organized in the attribution space. Interactive tabular display (d) populates when observations are selected. Contribution of the 1D basis affecting the horizontal position (e) parallel coordinate display of the variable attribution from all observations, and horizontal bars show the contribution to the current basis. Regression projection (f) uses the same horizontal projection and fixes the vertical positions to the observed $y$ and residuals (middle and right).}\label{fig:regressioncase}
\end{figure}
\end{CodeChunk}

@@

A tooltip displays the observation number/name and classification
information while the cursor hovers over a point. Linked brushing allows
the selection of points (left click and drag) where those points will be
highlighted across plots (Figure \ref{fig:classificationcase}a \& b).
The information corresponding to the selected points is populated on a
dynamic table (Figure \ref{fig:classificationcase}d). These interactions
aid the exploration of the spaces and, finally, the identification of
primary and comparison observations.

@@

\begin{itemize}
\tightlist
\item
  \textbf{Data:} predictors and response are unscaled complete numerical
  matrix. Most models and local explanations are scale-invariant. Keep
  the normality assumptions of the model in mind.
\item
  \textbf{Model:} any model and compatible explanation could be explored
  with this method. Currently, random forest models are applied via the
  package \textbf{randomForest} \citep{liaw_classification_2002},
  compatibility tree SHAP. Modest hyperparameters are used, namely: 125
  trees, the number of variables at each split, mtry = \(\sqrt{p}\) or
  \(p/3\) for classification and regression, and minimum size of
  terminal nodes \(max(1, n/500)\) or \(max(5, n/500)\) for
  classification and regression.
\item
  \textbf{Local explanation:} Tree SHAP is calculated for \emph{each}
  observation using the package \textbf{treeshap}
  \citep{kominsarczyk_treeshap_2021}. We opt to find the attribution of
  each observation in the training data and not fit to fit variable
  interactions.
\item
  \textbf{Cheem viewer:} after the model and full explanation space are
  calculated, each variable is scaled by standard deviations away from
  the mean to achieve common support for visuals. Statistics for mapping
  to color are computed on the scaled spaces.
\end{itemize}

The time to preprocess the data will vary significantly with the
complexity of the model and the LE. For reference, the FIFA data
contained 5000 observations of nine explanatory variables that took 2.5
seconds to fit a random forest model of modest hyperparameters.
Extracting the tree SHAP values of each observation took 270 seconds in
total. PCA and statistics of the variables and attributions took 2.8
seconds. These run times were from a non-parallelized session on a
modern laptop, but suffice it to say that most of the time will be spent
on the LVA. An increase in model complexity or data dimensionality will
quickly become an obstacle. Its reduced computational complexity makes
tree SHAP an excellent candidate to start. Alternatively, some package
and methods use approximate calculations of LEs, such as
\textbf{fastshap} \citet{greenwell_fastshap_2020}.

@@

There is more separation visible in the attribution space than in the
data space, as would be expected. The predicted vs observed plot reveals
a handful of misclassified observations. A Gentoo which has been wrongly
labeled as a Chinstrap is selected for illustration. The PI is a
misclassified point (represented by the asterisk in the global view and
a dashed vertical line in the tour view). The CI is a correctly
classified point (represented by an \(\times\) and a vertical dotted
line).

The radial tour starts from the attribution projection of the
misclassified observation (b, left). The important variables identified
by SHAP in the (wrong) prediction for this observation are mostly
\texttt{bl} and \texttt{bd} with small contributions of \texttt{fl} and
\texttt{bm}. This projection is a view where the Gentoo (purple) looks
much more likely for this observation than Chinstrap. That is, this
combination of variables is not particularly useful because the PI looks
very much like other Gentoo penguins. The radial tour is used to vary
the contribution of flipper length (\texttt{fl}) to explore this. (In
our exploration, this was the third variable explored. It is typically
helpful to explore the variables with more significant contributions,
here \texttt{bl} and \texttt{bd}. Still, when doing this, nothing was
revealed about how the PI differed from other Gentoos). On varying
\texttt{fl}, as it contributes increasingly to the projection (b,
right), more and more, this penguin looks like a Chinstrap. This
suggests that \texttt{fl} should be considered an important variable for
explaining the (wrong) prediction.

@@

\caption[Examining the LVA for a PI which is dark (orange) chocolate incorrectly predicted to be milk (green)]{Examining the LVA for a PI which is dark (orange) chocolate incorrectly predicted to be milk (green). From the attribution projection, this chocolate correctly looks more like dark than milk, which suggests that the LVA does not help understand the prediction for this observation. So, the contribution of Sugar is varied---reducing it corresponds primarily with an increased magnitude from Fiber. When Sugar is zero, Fiber contributes strongly toward the left. In this view, the PI is closer to the bulk of the milk chocolates, suggesting that the prediction put a lot of importance on Fiber. This chocolate is a rare dark chocolate without any Fiber leading to it being mistaken for a milk chocolate. (A video of the tour animation can be found at https://vimeo.com/666431143.)}\label{fig:casechocolates}
\end{figure}
\end{CodeChunk}

@@

Figure \ref{fig:caseames} selects the house sale 74, a sizable
under-prediction with an enormous Lot Area contribution. The CI has a
similar predicted price though the prediction was accurate and gives
almost no attribution to lot size. The attribution projection places
observations with high Living Areas to the right. The contribution of
Living Area contrasts the contribution of this variable. As the
contribution of Lot Area decreases, the predictive power decreases for
the PI, while the CI remains stationary. This large importance in the
Living Area is relatively uncommon. Boosting tree models may be more
resilient to such an under-prediction as they would up-weighting this
residual and force its inclusion in the final model.

@@

To address this challenge this paper provides a technique that builds on
LEs to explore the variable importance local to an observation. The LVA
is converted into an attribution projection from which variable
contributions are varied using a radial tour. Several diagnostic plots
are provided to assist with understanding the sensitivity of the
prediction to particular variables. A global view shows the data space,
explanation space, and residual plot. The user can interactively select
observations to compare, contrast, and study further. Then the radial
tour is used to explore the variable sensitivity identified by the
attribution projection.

This approach has been illustrated using four data examples of random
forest models with the tree SHAP LVA. LEs focus on the model fit and
help to dissect which variables are most responsible for the fitted
value. They can also form the basis of learning how the model has got it
wrong, when the observation is misclassified or has a large residual.
